Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life

### 2020-07-30 render2-mee.tex

\PassOptionsToPackage{unicode=true}{hyperref} % options for packages loaded elsewhere
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[]{article}
\usepackage{lmodern}
\usepackage{setspace}
\setstretch{2}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\usepackage{fixltx2e} % provides \textsubscript
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provides euro and other symbols
\else % if luatex or xelatex
  \usepackage{unicode-math}
  \defaultfontfeatures{Ligatures=TeX,Scale=MatchLowercase}
\fi
% use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
% use microtype if available
\IfFileExists{microtype.sty}{%
\usepackage[]{microtype}
\UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\IfFileExists{parskip.sty}{%
\usepackage{parskip}
}{% else
\setlength{\parindent}{0pt}
\setlength{\parskip}{6pt plus 2pt minus 1pt}
}
\usepackage{hyperref}
\hypersetup{
            pdftitle={Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life},
            pdfborder={0 0 0},
            breaklinks=true}
\urlstyle{same}  % don't use monospace font for urls
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Fix footnotes in tables (requires footnote package)
\IfFileExists{footnote.sty}{\usepackage{footnote}\makesavenoteenv{longtable}}{}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
\setlength{\emergencystretch}{3em}  % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
% Redefines (sub)paragraphs to behave more like sections
\ifx\paragraph\undefined\else
\let\oldparagraph\paragraph
\renewcommand{\paragraph}[1]{\oldparagraph{#1}\mbox{}}
\fi
\ifx\subparagraph\undefined\else
\let\oldsubparagraph\subparagraph
\renewcommand{\subparagraph}[1]{\oldsubparagraph{#1}\mbox{}}
\fi

% set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother

\usepackage{color}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{blindtext}
\usepackage{url}
\usepackage[left]{lineno}
\linenumbers

@@

\textbf{1. Luna L. Sanchez Reyes}

@@

Department of Botany, Faculty of Science, Charles University, Prague, Czech Republic

@@

\textbf{Word count}: 3722

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automatize
  some parts of the process of phylogenetic reconstruction (mainly character matrix construction)
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and choosing of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is an open‐source, command-line Python program that automatizes the update of published
  phylogenies by enriching underlying gene alignments with public DNA sequence data, and linking taxonomic information across databases.
  This provides a framework for comparison of published phylogenies with their updated versions, by using the conflict Application Programming Interface (API) function of the Open Tree of Life project.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on already available expert phylogenetic knowledge.
  Phylogeneticists and group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  We hope that the Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging more researchers to strive for better sharing practices. Physcraper can be used with any OS. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io/en/tutorial/index.html}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases provide an amazing resource for evolutionary estimation, but a large portion of sequence data available has never been incorporated into any phylogenetic estimate.

GenBank, the USA National Center for Biodiversity Information (NCBI) molecular database, release number 159 (April 15, 2007) hosts 72 million DNA sequences that are estimated to have the potential to resolve evolutionary relationships of most of its 241 000 distinct taxa (about 98.05\% of taxa in the NCBI taxonomy release 159; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}).
Yet, current public estimates of phylogenetic relationships are available for only about 100 000 taxa (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Hinchliff \emph{et al.} \protect\hyperlink{ref-hinchliff2015synthesis}{2015}; OpenTreeOfLife \emph{et al.} \protect\hyperlink{ref-opentreeoflife2019synth}{2019}), representing less than half of the taxonomic diversity with phylogenetically informative sequence data available in GenBank more than a decade ago.

Some of the discrepancy between data availability and phylogenetic estimates can be explained by the many phylogenies that are generated and published and not shared publicly in an accesible way (Drew \emph{et al.} \protect\hyperlink{ref-drew2013lost}{2013}; Magee \emph{et al.} \protect\hyperlink{ref-magee2014dawn}{2014}; McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2018bioessay}{2017}). However, there is also a lag between the amount of new DNA data generated and the analysis of these data in a phylogenetic context. Connecting new sequence data with existing estimates of evolutionary relationships is a challenge.
WHY IS IT A CHALLENGE

We address this gap by extending existing phylogenetic estimates with publicly available sequence data. By using a starting tree and single locus alignment, Physcraper, takes advantage of existing research, and extends trees using loci that taxon specialists have assessed and deemed appropriate for the phylogenetic scope.
The sequences added in the search are limited to a user specified taxon or monophyletic group, or within the taxonomic scope of the in-group of the starting tree.
These automated trees can provide a quick inference of potential relationships, of problems in the taxonomic assignments of sequences, and flag areas of potential systematic interest.

Physcraper leverages phylogentic data stored in Open Tree of Life and in TreeBase.
The Open Tree of Life (\url{https://opentreeoflife.github.io/}) is a project that unites phylogenetic inferences and taxonomy to provide a synthetic estimate of species relationships across the entire tree of life.
Open Tree of Life aims to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data.
Currently the tree comprises 2.3 million tips, of which around 90,000 of those taxa are represented by phylogenetic estimates - the rest are placed in the tree based on their taxonomic names.

The Open Tree of Life data store, \href{https://academic.oup.com/bioinformatics/article/31/17/2794/183373}{Phylesystem}, contains more than 4,500 phylogenetic trees from published studies.
The tips in these trees are mapped a unified taxonomy, which makes these data searchable in a phylogenetically explicit way.
This provides a resource for finding existing estimates of phylogenetic relationships,
and assessing which regions of the tree of life which are lacking available phylogenetic estimates.

By linking molecular data, available from databases such as the GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), to alignments and phylogenies, available in
the TreeBASE repository (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) and the Open Tree of Life (OpenTree) datastore (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}), we can place new biological data in an evolutionary context.

While the focus on single locus and gene sequence alignments could appear backwards-looking, in the age of genomics, single locus data has a lot to offer phylogenetics.
One major challenge of inferring phylogenies from genome scale data is inference of homology, and acquiring homologous data across divergent species.
Different research questions call for on different approaches to genomic sequencing, from whole genomes, to transcriptomes, to RadSeq and UCE's.
This variety of approaches results in non-overlapping data sets across taxa.
Even when the same sequencing approach is applied, such as RadSeq phylogenetic distance can cause allelic dropout at deeper divergences (Eaton \emph{et al.} \protect\hyperlink{ref-eaton2016misconceptions}{2016})
In contrast, single locus sequencing generates homologous data across large phylogenetic scales.

Indeed, some systematics support a classic phylogenetics approach (few markers thoughtfully curated) over the genomics approach (a massive amount of DNA markers that will overcome potential errors in the alignment coming from a lack of human curation).
Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are considered the gold standard for inferring species relationships {[}Song \emph{et al.} (\protect\hyperlink{ref-song2012resolving}{2012}); ROJAS ET AL. bats paper, take citations from there{]}.
It has also been suggested that manual curation of locus alignments produces better phylogenetic reconstructions and this has been demonstrated for genomic alignments (Fragoso-Martínez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the best of two worlds (massive amounts of newly released molecular data AND fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). This expert-curated alignments can be continuously enriched and updated by incorporating newly released data from public molecular databases.

In leveraging existing homology statements in the form of alignments, this approach differs from existing approaches that automatize the assembly of DNA alignments from the GenBank database for phylogenetic reconstruction (``phylogenetic pipelines'') such as PHYLOTA (Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}), PHLAWD (Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}), and SUPERSMART (Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
Physcraper shares a similar conceptual framework to Pumper (Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo-carrasco_pumper:_2014}{2014}), but that software is not currently supported or developed (\emph{or runnable at all honestly\ldots{}})

Data input availability:
As of April 2014, the TreeBASE repository hosted about 8 200 curated alignments, providing information on evolutionary relationships of around 100 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}.
This database provides an untapped source of valuable expert knowledge with the potential to update phylogenetic relationships in several different regions of the tree of life.

The Phylesystem (OpenTree's datastore) (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}) automatically incorporates phylogenies from TreeBASE, and saves metadata linking the original tree to its corresponding alignment repository in TreeBASE. If there are multiple alignments, TreeBASE does not always indicate how they were used to generate the tree. This provides a loose means of linking the tree with the exact alignment that generated it.

Often, linking data in an original alignment with its corresponding phylogeny has to be done by a human curator.
Moreover, different data repositories follow different systems for taxon and study identification, posing a real challenge to automatically link data from across databases that belong to the same taxon and study.
OpenTree's metadata system incorporates taxon identifiers from a variety of taxonomies and repositories, including the NCBI taxonomy, GBIF, etc., providing a way to automatically link data from different databases.

Physcraper is a Python encoded pipeline that relies on the OpenTree metadata system to connect databases through unique taxon identifiers.
main functionality is to connect phylogenies stored in the OpenTree phylesystem, with alignments from TreeBASE and newly released DNA data from GenBank, in order to update previously known phylogenetic relationships in a continuous manner.
EXPLIAN WHY PYTHON IS BEST FOR THIS.
By design, this approach focusses on interoperability. By automating taxonomic name matching across NCBI, OpenTree, GBIF and other data stores, users can easily perform downstream analyses.
For example, it allows automatizing and standardizing the comparison of phylogenetic hypotheses with currently known relationships in the Open Tree of Life taxonomy and phylogenetic data store.

We envision Physraper as a staple tool to make data connections across databases in a phylogenetic context for the advantage of phylogenetics and biology in general, as well as an effort towards fully reproducible workflows in phylogenetics.

\hypertarget{how-does-physcraper-work}{%
\section{How does Physcraper work?}\label{how-does-physcraper-work}}

@@

\caption{Physcraper general framework.}\label{fig:framework}
\end{figure}

The Physcraper software is fully described on its documentation website at \url{https://physcraper.readthedocs.io/en/latest/}, along with installation, tutorials, examples and function usage.
The general Physcraper workflow is shown in Figure \ref{fig:framework}. Next, we will describe the technical details of each step of the workflow.

\hypertarget{the-input-a-tree-and-an-alignment}{%
\subsection{The input: a tree and an alignment}\label{the-input-a-tree-and-an-alignment}}

\begin{itemize}
\tightlist
\item
  In order to take advantage of the OpenTree tools, the input tree needs to be stored in the
  OpenTree phylesystem \href{https://github.com/opentreeoflife/phylesystem}{github.com/opentreeoflife/phylesystem}, or need to have the tip labels mapped to the unified OpenTree taxonomy via bulk taxonomic name resolution (\url{https://tree.opentreeoflife.org/curator/tnrs/}).
  The main
  reason for this is that trees in phylesystem have a set of user defined characteristics
  that are essential for automatizing the phylogeny update process. The most relevant of these being the standardization of taxonomic names and the definition of ingroup and outgroup. Outgroup and ingroup taxa in the original tree are identified and tagged. This allows to automatically set the root for the updated tree on the final steps of the pipeline.
  Currently, only trees connected to
  a published study can be stored in the phylesystem (although there are plans to
  allow storage of unpublished trees).
  A user can choose from among the 1216 published trees supporting the resolved nodes of the synthetic tree in the OpenTree website (\href{https://tree.opentreeoflife.org/about/synthesis-release/v12.3}{See OpenTree's website about}). If the published tree you are interested in updating is not in there, you can upload it via OpenTree's curator tool (\href{https://tree.opentreeoflife.org/curator}{www.opentreeoflife.org/curator}.
  If the user is not ready to make the input tree public, \texttt{physcraper} can use as input a tree, and aliognment, and a file of mappings between the tip labels and taxonomic names.
\item
  The input alignment should be a single locus alignment that was used to generate the tree. For public data, the original
  alignments are often stored in a public repository such as TreeBase (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Vos \emph{et al.} \protect\hyperlink{ref-vos2012nexml}{2012}),
  DRYAD (\href{http://datadryad.org/}{www.datadryad.org}), or the journal were the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper can download it directly,
  either from the TreeBASE website (\href{https://treebase.org/}{www.treebase.org})
  or through the TreeBASE GitHub repository (SuperTreeBASE; \href{https://github.com/TreeBASE/supertreebase}{github.com/TreeBASE/supertreebase}).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
\item
  A taxon name matching step is performed to verify that all taxon names on the tips
  of the tree are in the DNA character matrix and vice versa.
\item
  A ``.csv'' file with the summary of taxon name matching is produced for the user.
\item
  Unmatched taxon names are dropped from both the tree and alignment.
  Technically, just one matching name is needed to perform the searches. Please, see next section.
\item
  A ``.tre'' file and a ``.fas'' file containing only the matched taxa are generated and saved in the \texttt{inputs} folder to be used in the following steps.
\end{itemize}

@@

\begin{itemize}
\item
  Technically, any DNA molecular database can be used to search for new sequences.
  By default we rely on the GenBank database.
  The new sequence search can be performed on the remote database or in a local database.
\item
  The next step is to identify the search taxon within the reference taxonomy. The search taxon will be used to constraint the DNA sequence search on the nucleotide database within that taxonomic group. Because we are using the NCBI nucleotide database, by default the reference taxonomy is the NCBI taxonomy. The search taxon can be determined by the user. If none is provided, then the search taxon is identified as the Most Recent Common Ancestor (MRCA) of the
  matched taxa belonging to the ingroup in the OPenTree synthetic tree, that is also a named clade in the reference taxonomy. FIGURE RECOMMENDED. This is known as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred in the literature as the Least Inclusive Common Ancestral Taxon - LICA) CITATION NEEDED.
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the reference taxonomy.
  To automatically identify the MRCAT of a group of taxon names, we make use of the OpenTree taxonomic tool \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{Taxonomy-API-v3} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}).

@@

  The \texttt{blastn} function from the BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) is used for local-database sequence searches.
  For remote-database searches, we modified the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function in the \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} to accept an alternate BLAST URL. This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST without NCBI's required wait times.

@@

  A pairwise BLAST search is performed. This means that each sequence
  in the alignment is BLASTed against DNA sequences in a nucleotide database constrained to the search
  taxon. Results from each one of these BLAST runs are recorded, and matched sequences are saved
  along with their corresponding identification numbers (accession numbers in the case of the GenBank database). This information will be used later to store the whole sequences in a dedicated library within the ``physcraper'' folder, allowing for secondary analyses to run significantly faster.

@@

  Reverse complement sequences are identified and translated.

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{setspace}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Physcraper: a python package for continual update of evolutionary estimates using the Open Tree of Life},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin = 1in]{geometry}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{color}
\usepackage{hyperref}
\usepackage{caption}
\usepackage{blindtext}
\usepackage{url}
\usepackage[left]{lineno}
\linenumbers

@@

\setstretch{2}
\textbf{1. Luna L. Sanchez Reyes}

@@

School of Natural Sciences, University of California, Merced
Department of Botany, Faculty of Science, Charles University, Prague, Czech Republic

@@

@@

\begin{enumerate}
\def\labelenumi{\arabic{enumi}.}
\item
  Phylogenies are a key part of research in many areas of biology. Tools that automatize
  some parts of the process of phylogenetic reconstruction, mainly molecular character matrix assembly,
  have been developed for the advantage of both specialists in the field of phylogenetics and nonspecialists.
  However, interpretation of results, comparison with previously available phylogenetic
  hypotheses, and choosing of one phylogeny for downstream analyses and discussion still impose difficulties
  to one that is not a specialist either on phylogenetic methods or on a particular group of study.
\item
  Physcraper is a command-line Python program that automatizes the update of published
  phylogenies by enriching underlying gene alignments with public DNA sequence data. It builds upon tools from the Open Tree of Life project to link taxonomic information across databases, providing a framework for straightforward comparison of published phylogenies with their updated versions.
\item
  Physcraper can be used by the nonspecialist, as a tool to generate phylogenetic
  hypotheses based on publicly available expert phylogenetic knowledge.
  Phylogeneticists and group specialists will find it useful as a tool to facilitate molecular dataset gathering and comparison
  of alternative phylogenetic hypotheses (topologies).
\item
  We hope that the Physcraper workflow demonstrates the benefits of doing open science for phylogenetics, encouraging more researchers to strive for better sharing practices. Physcraper can be used with any OS and is released under an open-source license. Detailed instructions for installation and
  use are available at \href{https://physcraper.readthedocs.io/en/tutorial/index.html}{https://physcraper.readthedocs}.
\end{enumerate}

@@

Phylogenetic estimates of evolutionary relationships capture the shared history of living organisms, and provide key context for all our biological observations.
Public biological databases constitute an amazing resource for evolutionary estimation, but a large portion of molecular data publicly available has never been incorporated into any phylogenetic estimate. Extending existing phylogenetic estimates with new DNA sequence data, geographical location, fossil data, etc., in a reproducible and continuous manner is possible by automatizing connections between biological databases. Here, we propose Physcraper, a tool that uses existing phylogenetic research and data from public biological databases to update a starting tree and single locus alignments, to build upon molecular data that taxon specialists have assessed and deemed appropriate for a specific phylogenetic scope.

The prevalence of taxonomic idiosincracies across databases represent a big challenge to automatically connect data from disparate biological databases in a phylogenetic context. To standardize taxonomic names, a unified taxonomy is needed. The main aim of the Open Tree of Life (OpenTree from now on) project
is to construct a comprehensive, dynamic and digitally-available tree of life by synthesizing published phylogenetic trees along with taxonomic data. Currently, OpenTree's ``synthetic'' tree comprises 2.3 million tips, of which around 90,000 are represented by phylogenetic estimates - the remaining 1.4 million taxa are placed in the tree based on their taxonomic assignment. To achieve this, OpenTree unifies taxonomy from various databases (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}) such as the USA National Center for Biodiversity Information (NCBI) molecular database GenBank (Benson \emph{et al.} \protect\hyperlink{ref-benson2000genbank}{2000}; Wheeler \emph{et al.} \protect\hyperlink{ref-wheeler2000database}{2000}), the Global Biodiversity Information Facility (GBIF; Secretariat \protect\hyperlink{ref-secretariat2017gbif}{2017}), and the World Register of Marine Species {[}WoRMS; www.marinespecies.org/{]}, providing a key resource that can be used to connect data from virtually any biological database to phylogenetic data that has been standardized to OpenTree's unified taxonomy.

Another challenge to incorporating molecular data from public databases to update phylogenetic knowledge is assembling high-quality homology hypotheses. Species tree reconstructions from multi-gene data sets taking into account the multispecies coalescent model are seen as the gold standard for inferring species relationships (Song \emph{et al.} \protect\hyperlink{ref-song2012resolving}{2012}).
While genomics have allowed us to obtain large quantities of molecular data from one organism, different research questions call for different genomic sequencing approaches, from whole genomes, to transcriptomes, restriction-site associated DNA sequencing (RADSeq), single nucleotide polymorphisms (SNPs), microsatellites, and ultra-conserved elements (UCEs).
Having a variety of approaches has resulted in largely non-overlapping genomic data sets across taxa. Moreover, some types of genomic data are not suitable for phylogenetic inference at any scale. For example, phylogenetic distance can cause allelic dropout at deeper divergences with RADSeq data sets (Eaton \emph{et al.} \protect\hyperlink{ref-eaton2016misconceptions}{2016}).
In contrast, decades of single locus sequencing have generated homologous DNA data
that can be used for phylogenetic reconstruction at many scales. Indeed, more than a decade ago, GenBank release number 159 (April 15, 2007) already hosted 72 million non-genomic DNA sequences that were gauged to have the potential to resolve phylogenetic relationships of most of the 241, 000 distinct taxa in the NCBI taxonomy at the time (about 98.05\% of taxa; Sanderson \emph{et al.} \protect\hyperlink{ref-sanderson2008phylota}{2008}). Assembling a DNA alignment from such a massive database can be done ``by hand'', but it requires huge amounts of time and it is mostly a non-reproducible method. Computer pipelines that make DNA sequence search faster and more efficient, as well as more reproducible have been utilized to study evolutionary relationships among a variety of organisms (e.g., Smith \emph{et al.} \protect\hyperlink{ref-smith2009mega}{2009}; Izquierdo-Carrasco \emph{et al.} \protect\hyperlink{ref-izquierdo2014pumper}{2014}; Antonelli \emph{et al.} \protect\hyperlink{ref-antonelli2017toward}{2017}).
Still, there is a large tradition in phylogenetics of building molecular character matrices of few, thoughtfully curated markers, as manual curation of alignments often improves phylogenetic reconstructions (Fragoso-Martı́nez \emph{et al.} \protect\hyperlink{ref-fragoso2017pilot}{2017}).

A way to incorporate the best of two worlds (massive amounts of newly released molecular data and fine-grained curation from human experts) is to rely on published manually curated homology hypotheses as ``jump-start'' alignments (Morrison \protect\hyperlink{ref-morrison2006multiple}{2006}). The TreeBASE database (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}) hosts about 8, 200 publicly accesible alignments, providing information on evolutionary relationships of around 100, 000 distinct taxa \href{https://www.treebase.org/treebase-web/home.html\#:~:text=TreeBASE\%20is\%20produced\%20and\%20governed,mapped\%20to\%20104\%2C593\%20distinct\%20taxa.}{(see TreeBASE's website about)}, representing an untapped source of valuable expert knowledge with the potential to continuously enrich and update phylogenetic relationships in several different regions of the tree of life, by incorporating data from molecular databases that has not been included in any public phylogenetic estimate. The phylogenies associated to all these alignments have been integrated to the OpenTree's datastore, the Phylesystem (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}), and metadata linking these trees to their corresponding alignment repository in TreeBASE is available.

We propose Physcraper as a Python encoded pipeline that relies on programmatic access protocols to link molecular data from GenBank, to alignments from TreeBASE and phylogenies from OpenTree's Phylesystem, to automatically update and enrich phylogenetic knowledge based on expertly-curated homology hypotheses. Physcraper also provides new types of access to various general OpenTree programmatic tools for custom comparison of phylogenetic hypotheses.
It focuses on database interoperability, by automatizing taxonomic name matching across biological databases, so users can perform any downstream analysis straightforwardly, as well as placing new data from virtually any biological database in an evolutionary context.

\hypertarget{the-physcraper-framework}{%
\section{The Physcraper framework}\label{the-physcraper-framework}}

@@

\caption{The Physcraper software is fully described on its documentation website at physcraper.readthedocs.io, along with installation instructions, tutorials, examples and function usage documentation.}\label{fig:framework}
\end{figure}

The general Physcraper framework is shown in Figure \ref{fig:framework}. Next, we will describe the technical details of each step of the workflow.

\hypertarget{the-inputs-a-tree-and-an-alignment}{%
\subsection{The inputs: a tree and an alignment}\label{the-inputs-a-tree-and-an-alignment}}

\begin{itemize}
\item
  In order to take advantage of the OpenTree tools, it is reccommended that the input tree is either stored in the
  OpenTree \href{https://github.com/opentreeoflife/phylesystem}{Phylesystem}, or submitted via OpenTree's curator \href{https://tree.opentreeoflife.org/curator}{application} (McTavish \emph{et al.} \protect\hyperlink{ref-mctavish2015phylesystem}{2015}). If the user is not ready to make the input tree public, tree tip labels must be standardized to the unified OpenTree taxonomy using the bulk Taxonomic Name Resolution Service \href{https://tree.opentreeoflife.org/curator/tnrs/}{TNRS} tool. \emph{Should we saythat it is based on Boyle's 2012 algorithm? or is it a TNRS algorithm independently developed by the OpenTree team?}
  This step is referred to as taxon name mapping. If taxonomic names can't be mapped, their taxonomic information will be excluded from further analysis.
  Mapping tip names to OpenTree's unified taxonomy saves a set of user defined characteristics
  that are essential for automatizing the phylogeny updating process. The most relevant of these is the taxonomic name mapping and the definition of ingroup and outgroup taxa, allowing to automatically set the root for the updated tree on the final steps of the pipeline. Currently, only trees connected to a published study can be stored in the Phylesystem.
  Users can choose from among the 2, 950 studies in OpenTree's Phylesystem that have alignments on
  TreeBASE.
\item
  The input alignment should be a single locus alignment that was used to generate the tree. Alignments are often stored in a public repository such as TreeBase (Piel \emph{et al.} \protect\hyperlink{ref-piel2009treebase}{2009}; Vos \emph{et al.} \protect\hyperlink{ref-vos2012nexml}{2012}),
  DRYAD (\href{http://datadryad.org/}{www.datadryad.org}), or a data repository associated with the journal where the tree was originally published.
  If the alignment is stored in TreeBase, Physcraper downloads it directly,
  either from the TreeBASE website (\href{https://treebase.org/}{www.treebase.org})
  or through the TreeBASE GitHub repository (SuperTreeBASE; \href{https://github.com/TreeBASE/supertreebase}{github.com/TreeBASE/supertreebase}).
  If the alignment is on another repository, or constitutes personal data, a path to a local copy of the alignment has to be provided.
\item
  Single locus alignments sometimes have fewer taxa than the tree inferred from the full concatenated data, simply because a single molecular marker usually does not cover all the taxa sampled for the full phylogenetic analysis. Physcraper prunes the input tree to taxa found in the alignment, and verifies that all taxon names on the tips of the tree are in the DNA character matrix and vice versa. Technically, just one taxon name (and its corresponding sequence in the alignment) is needed to continue the algorithm. See next section.
\item
  A summary ``csv'' file with results from the taxon name standardization and pruning is produced for the user.
\item
  The standardized and pruned tree and alignment (checked tree and alignment from now on) are output as ``newick'' and ``fasta'' respectively in the ``inputs'' folder to be used in the following steps.
\end{itemize}

@@

\begin{itemize}
\item
  Physcraper uses the GenBank DNA database as source to search for new sequences. The DNA sequence search can be performed on the GenBank remote database or in a GenBank local database set up by the user, which can speed up the search process. Detailed instructions to setup a local database are provided on Physcraper's software documentation.
\item
  The next step is to identify a ``search taxon'' to constrain the sequence search on the GenBank database within that taxonomic group.
  The search taxon can be chosen by the user from the NCBI taxonomy.
  If none is provided, then the search taxon is automatically set using the taxa in the input tree labeled as the ``ingroup'' (Fig. \ref{fig:framework}).
  The search taxon is The Most Recent Common Ancestor (MRCA) of the ingroup taxa in the OpenTree synthetic tree, that is also a named clade in the NCBI taxonomy.
  This is known in the OpenTree as the Most Recent Common
  Ancestral Taxon (MRCAT; also referred as the Least Inclusive Common Ancestral taxon - LICA).
  The MRCAT can be different from the phylogenetic MRCA when the latter is an unnamed clade in the synthetic tree.
  To identify the MRCAT of a group of taxon names, we use the OpenTree \href{https://github.com/OpenTreeOfLife/germinator/wiki/Taxonomy-API-v3\#mrca}{taxonomic tool v3} (Rees \& Cranston \protect\hyperlink{ref-rees2017automated}{2017}).

@@

  The \texttt{blastn} function from the BLAST command line tools (Camacho \emph{et al.} \protect\hyperlink{ref-camacho2009blast}{2009}) is used for local database sequence searches.
  For remote database searches, we modified the BioPython (Cock \emph{et al.} \protect\hyperlink{ref-cock2009biopython}{2009}) BLAST function from the \href{https://biopython.org/DIST/docs/api/Bio.Blast.NCBIWWW-module.html}{NCBIWWW module} to accept an alternative BLAST address (URL). This is useful when a user has no access to the computer capacity needed to setup a local database, and a local blast database can be set up on a remote machine to BLAST avoiding NCBI's required waiting times, which slow down the searches markedly.

@@

  A constrained BLAST search is performed, in which each sequence
  in the alignment is BLASTed once against all database DNA sequences belonging to the search
  taxon. All results from each BLAST run are stored, and sequences with match scores better than the e-value cutoff (default to 0.00001) are saved
  along with their corresponding metadata, i.e., their GenBank accession number.
  The full sequence for each match is downloaded from NCBI into a dedicated library within the ``physcraper'' folder, allowing for secondary analyses to run significantly faster.

@@

  All BLAST and filtering parameters can be customized by the user.
